---
title: "io102-marginal_effects"
author: "Peter Higgins"
date: "2024-12-18"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Marginal Effects

What are marginal effects? How does the {marginaleffects} package help?

```{r marginal}
library(marginaleffects)
library(tidyverse)
library(broom.helpers)

mod <- lm(mpg ~ vs * hp * am, data = mtcars)
tbl <- avg_comparisons(mod, variables = "hp", by = c("vs", "am"))
tidy_tbl <- tidy_avg_comparisons(mod, variables = "hp", by = c("vs", "am"))
tbl

View(tbl)

tidy_tbl

```

## Main Functions

| Goal                         | Function           |
|------------------------------|--------------------|
| Predictions                  | predictions()      |
|                              | avg_predictions()  |
|                              | plot_predictions() |
| Comparisons                  | comparisons()      |
|                              | avg_comparisons()  |
|                              | plot_comparisons() |
| Slopes                       | slopes()           |
|                              | avg_slopes()       |
|                              | plot_slopes()      |
| Grids                        | datagrid()         |
| Hypotheses and Equivalence   | hypotheses()       |
| Bayes, Bootstrap, Simulation | get_draws()        |
|                              | inferences()       |
|                              |                    |

## Including Plots

More examples of marginal effects

```{r plots}
plot(pressure)
```

Reference book in Marginal Effects Zoo <https://marginaleffects.com/>
