---
title: "R Notebook"
output: html_notebook
editor_options: 
  chunk_output_type: inline
---

This is an example of use of the blockrand package to create a randomization of two assignments with 5 strata in a randomized permuted block design

```{r setup}
library(blockrand)
library(knitr)
library(kableExtra)
library(tidyverse, quietly = T)
```

Our goal is to have 500 assignments available across 5 strata.

```{r}
set.seed(1234)
stratum1 <- blockrand(n = 100,
                      block.prefix = "B_",
                      stratum = "Stratum_1",
                      num.levels = 2,
                      levels = c("Patient Navigator", "Control")
                      ) %>% filter(id <101)
stratum1
```

Now stratum 2

```{r}
set.seed(0714)
stratum2 <- blockrand(n = 100,
                      block.prefix = "B_",
                      stratum = "Stratum_2",
                      num.levels = 2,
                      levels = c("Patient Navigator", "Control")
                      ) %>% filter(id <101)
stratum2
```


Now stratum 3

```{r}
set.seed(1965)
stratum3 <- blockrand(n = 100,
                      block.prefix = "B_",
                      stratum = "Stratum_3",
                      num.levels = 2,
                      levels = c("Patient Navigator", "Control")
                      ) %>% filter(id <101)
stratum3
```

Now stratum 4

```{r}
set.seed(314159)
stratum4 <- blockrand(n = 100,
                      block.prefix = "B_",
                      stratum = "Stratum_4",
                      num.levels = 2,
                      levels = c("Patient Navigator", "Control")
                      ) %>% filter(id <101)
stratum4
```

Now Stratum 5

```{r}
set.seed(2718)
stratum5 <- blockrand(n = 100,
                      block.prefix = "B_",
                      stratum = "Stratum_5",
                      num.levels = 2,
                      levels = c("Patient Navigator", "Control")
                      ) %>% filter(id <101)
stratum5
```

now bind these together in one assignment sheet

```{r}
assign500 <- rbind(stratum1, stratum2, stratum3, stratum4, stratum5)
assign500 %>% 
  rename(num = id) %>% 
  select(stratum, num, treatment, block.id, block.size) %>% 
  knitr::kable() %>% 
  kable_styling()
```



Add a new chunk by clicking the *Insert Chunk* button on the toolbar or by pressing *Cmd+Option+I*.

When you save the notebook, an HTML file containing the code and output will be saved alongside it (click the *Preview* button or press *Cmd+Shift+K* to preview the HTML file). 

The preview shows you a rendered HTML copy of the contents of the editor. Consequently, unlike *Knit*, *Preview* does not run any R code chunks. Instead, the output of the chunk when it was last run in the editor is displayed.

