---
title: "Comparing 2 groups"
author: "Peter Higgins"
date: "1/2/2019"
output: powerpoint_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(tidyverse)
library(broom)
```

## Comparing Two Measures of Centrality
- Means and Medians
- t tests and wilcoxon
- for numerical continuous data


## Common Problem
- Comparing two groups
  - Mean or median vs. expected
  - Two arms of study - independent
  - Pre and post / spouse and partner / left vs right arm – paired groups
- Are the means significantly different?
- Or the medians (if not normally distributed)?


## How Skewed is Too Skewed?
- Formal test of normality = Shapiro-Wilk test
- Use base data set called ToothGrowth


```{r data, message= FALSE,warning=TRUE, echo=TRUE}
library(tidyverse)
data <- ToothGrowth
head(data)
```

## Visualize the Distribution of data$len in ggplot
- Use geom_histogram or geom_density
- Bonus points: facet by supplement treatment
- Your turn to try it



```{r density}
data %>% 
ggplot(mapping = aes(len)) +
  #geom_histogram() +
  geom_density() + 
  facet_wrap(~supp)
```

```{r histogram, echo=TRUE, warning=FALSE, message=FALSE}
data %>% 
ggplot(mapping = aes(len)) +
  geom_histogram() +
  #geom_density() + 
  facet_wrap(~supp)
```

## Visualize the Distribution of data$len in ggplot
- The OJ group is left skewed
- May be problematic for using means
- formally test with Shapiro-Wilk
```{r shapirowilk, echo=TRUE, warning=FALSE, message=FALSE}
data$len %>% 
shapiro.test()
```

## Results of Shapiro-Wilk
- p-value = 0.1091
- p not < 0.05
- Acceptably close to normal
- OK to compare means rather than medians
- can use t test rather than wilcoxon test
  - if p is < 0.05, use wilcoxon test
  - also known as Mann-Whitney test
  - a rank-based (non-parametric) test
  
## Try it yourself
- use df <- msleep
```{r shapirowilk2, echo=TRUE, warning=FALSE, message=FALSE}
df <- msleep 
head(df$sleep_total)
```
- test the normality of total sleep hours in mammals

## Mammal sleep hours
```{r shapirowilk3, echo=TRUE, warning=FALSE, message=FALSE}
shapiro.test(df$sleep_total)
```
- meets criteria - acceptable to consider normally distributed
- now consider - is the mean roughly 8 hours of sleep per day?

## One Sample T test
- univariate test
  - Ho: mean is 8 hours
  - Ha: mean is not 8 hours
- can use t test because shapiro.test is NS

## How to do One Sample T test
```{r univar T, eval=FALSE, echo=TRUE}
t.test(df$sleep_total, alternative = "two.sided",
       mu = 8)
```
- Try it out, see if you can interpret results

## Interpreting the One Sample T test
```{r univar T2, eval=TRUE, echo=FALSE}
t.test(df$sleep_total, alternative = "two.sided",
       mu = 8)
```
- p is highly significant
  - can reject the null, accept alternative
  - sample mean 10.43, CI 9.46-11.41

## What are the arguments of the t.test function?
- x = vector of continuous numerical data
- y= NULL - optional 2nd vector of continuous numerical data
- alternative = c("two.sided", "less", "greater"),
- mu = 0
- paired = FALSE
- var.equal = FALSE
- conf.level = 0.95
- [documentation](https://www.rdocumentation.org/packages/stats/versions/3.5.1/topics/t.test)



## Fine, but what about 2 groups?
- consider df$vore
```{r vore, eval=T, echo=T}
table(df$vore)
```
- hypothesis - herbivores need more time to get food, sleep less than carnivores
- how to test this?
  - normal, so can use t test for 2 groups
  
## Setting up 2 group t test
- formula interface: outcome ~ groupvar 
```{r vore2, eval=F, echo=T}
df %>% 
  filter(vore %in% c("herbi", "carni")) %>% 
  t.test(formula = sleep_total ~ vore, data = .)
```
- Try it yourself
- What do the results mean?

## Results of the 2 group t test
```{r vore3, eval=T, echo=F}
df %>% 
  filter(vore %in% c("herbi", "carni")) %>% 
  t.test(formula = sleep_total ~ vore, data = .)
```
## Interpreting the 2 group t test
- Welch t-test (not Student)
  - Welch does NOT assume equal variances in each group
- p value NS
- accept null hypothesis
  - Ho: means of groups roughly equal
  - Ha: means are different
  - 95% CI crosses 0
- Carnivores sleep a little more, but not a lot

## 2 group t test with wide data
- You want to compare column A with column B (data are not tidy)
- Do mammals spend more time awake than asleep?
```{r wide, eval=F, echo=T}
t.test(x = df$sleep_total, y = df$awake, data = msleep)
```

## Results of 2 group t test with wide data

```{r wide2, eval=T, echo=T}
t.test(x = df$sleep_total, y = df$awake, data = msleep)
```

## 3 Assumptions of Student's t test  
1. Sample is normally distributed (test with Shapiro)
2. Variances are homogeneous (homoskedasticity) (test with Levene)
3. Observations are independent
  + not paired like left vs. right colon
  + not paired like spouse and partner
  + not paired like measurements pre and post Rx
  
## Testing Assumptions of Student's t test  
- Normality - test with Shapiro
  - If not normal, Wilcoxon >  t test
- Equal Variances - test with Levene
  - If not equal, Welch t > Student's t
- Observations are independent
  - Think about data collection
  - are some observations correlated with some others?
  - If correlated, use paired t test
  
## Getting results out of t.test
- Use the tidy function from the broom package
- Do carnivores have bigger brains than insectivores?
```{r tidy, eval=F, echo=T}
df %>% 
  filter(vore %in% c("carni", "insecti")) %>% 
t.test(formula = brainwt ~ vore, data = .) %>% 
  tidy() ->
result
result
```

## Getting results out of t.test

```{r tidy7, eval=T, echo=F}
df %>% 
  filter(vore %in% c("carni", "insecti")) %>% 
t.test(formula = brainwt ~ vore, data = ., var.equal=TRUE) %>% 
  tidy() ->
result
result
```
## Reporting the results from t.test using inline code
- use backticks before and after, start with r
  - i.e. My result is [backtick]r code here[backtick].
- The mean brain weight for carnivores was `r result$estimate1`
- The mean brain weight for herbivores was `r result$estimate2`
  - The difference was `r result$estimate`
- The t statistic for this `r result$method` was `r result$statistic`
- The p value was `r result$p.value`
  - The confidence interval was from `r round(result$conf.low,2)` to `r round(result$conf.high,2)`


## For Next Time  
- Skewness and Kurtosis
- Review Normality
  - When to use Wilcoxon
- Levene test for equal variances
  - When to use Welch t vs. Student's t
- Paired t and Wilcoxon tests
