---
title: "Comparing 2 groups"
author: "Peter Higgins"
date: "1/2/2019"
output: powerpoint_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(tidyverse)
library(broom)
```

## Review from t test part 1
- Comparing Two Measures of Centrality
  - is it normal (Shapiro test)?
    - if yes, use t test
    - if no, use Wilcoxon rank test (nonparametric)
  - are the observations paired?
    - if yes, use paired Wilcoxon or paired t test
    - if no, can use usual versions
  - are the variances equal in 2 groups (Levene test)?
    - if yes, OK to use Student's t test
    - if no, must use Welch t test (unequal var)

## Review how to do t test part 1  
  - one sample
    - t.test(df$sleep_total, alternative = "two.sided",        mu = 8)
  - formula interface: outcome ~ groupvar
df %>%    filter(vore %in% c("herbi", "carni")) %>%    t.test(formula = sleep_total ~ vore, data = .)
  - use broom::tidy() to get results into a dataframe
  - can use results in text with inline code
  - use backticks before and after, start with r
i.e. My result is [backtick]r code here[backtick].


## What We have Left to Cover
- General info on Skewness and Kurtosis
- When to use t tests vs. rank (wilcoxon)
- Levene test for equal variances
  - When to use Welch t vs. Student’s t
- Paired t and Wilcoxon tests


## Skewness
-  A term for asymmetric data - obviously not normal
  - Left-skewed - long leftward tail
  -   aka negatively skewed
    - mean less than the median
    - typical of age at death in US, number of questions answered on a survey, sq_ft/$ in Ann Arbor housing
    - often has a natural upper bound
  - Right-skewed - long rightward tail
    - aka positively skewed
    - mean greater than the median
    - typical of length of stay, medical costs, children per family
    - mostly low, but some huge outliers
    - often has a natural lower bound
    
## Skewness example: Old Faithful geyser eruption duration
- Duration of eruptions in minutes is negatively skewed
```{r geys}
library(MASS)
#skewness(geyser$duration) 
hist(geyser$duration)
```

## Skewness example: Old Faithful eruption wait times
- Wait time for eruptions in minutes is negatively skewed
```{r geyser2}
library(MASS)
#skewness(geyser$waiting) 
hist(geyser$waiting)
```
## How Skewed is Too Skewed?
- Formal test of normality = Shapiro-Wilk test
- Use base data set called ToothGrowth
- test the len variable

```{r data2, message= FALSE,warning=TRUE, echo=TRUE}
library(tidyverse)
library(MASS)
shapiro.test(geyser$waiting)
```

## Kurtosis
- Descriptive of how tightly bunched data are
  - leptokurtotic - histogram has tall peak, little spread 
    - think leaping high = lepto
    - values <3 are platykurtotic
  - platykurtotic - histogram is mostly flat, wide spread
    - think flat, like a platypus beak
    - values <3 are platykurtotic
  - kurtosis values ~ 3 are mesokurtotic

```{r data, message= FALSE,warning=TRUE, echo=TRUE}
library(tidyverse)
library(MASS)
#kurtosis(geyser$waiting)
```
## Decision tree for Comparing 2 group data

1. Is it normal (Shapiro test)?
  - yes - t.test
  - no - wilcoxon
2. Is it paired data?
  - use paired variants of either t.test or wilcoxon
3. If normal, are the variances of the two groups equal (Levene test)?
  - if yes, use Student t test (var.equal=TRUE)
  - if no, use Welch t test (var.equal = FALSE)

## Tooth length data
- We know it is reasonably normal from Shapiro test
  - Ok to use t test
- Should we use the Student version (var.equal) or the Welch version?
  - Welch (var.equal = FALSE) is the default
- Determine this with Levene test for 2 groups (in car package)
  - use Bartlett test if >2 groups and normally distributed

```{r levene, message= FALSE,warning=TRUE, echo=TRUE}
library(tidyverse)
library(car)
data <- ToothGrowth
data %>% 
  filter(dose > 0.9) %>% 
  leveneTest(len ~ as.factor(dose), data = .)
# note has to be as.factor - will not evaluate if groupvar is numeric
```

## Interpretation of Levene's test

- the null hypothesis is that all populations variances are equal;
- the alternative hypothesis is that at least two of them differ.
```{r chunk2, ref.label="levene", echo=FALSE, eval=TRUE}
```

- will not reject null, OK to use Student t test

## Arguments for t.test
x = vector of continuous numerical data
alternative = c(“two.sided”, “less”, “greater”),
mu = 0
paired = FALSE
var.equal = FALSE
conf.level = 0.95
Open the t.test help - type in console "?t.test"


## Mammal sleep hours for t test

First with Student t test, compare herbivores to omnivores by sleep_rem
```{r t sleep2, echo=TRUE, warning=FALSE, message=FALSE}
df <- msleep
df %>% 
  filter(vore != 'carni') %>% 
  filter(vore != 'insecti') %>% 
  t.test(sleep_rem ~ vore, data=., var.equal=TRUE)
```

Now with Welch's t test, do the same thing
compare herbivores to omnivores by sleep_rem
```{r t sleep, echo=TRUE, warning=FALSE, message=FALSE}
df <- msleep
df %>% 
  filter(vore %in% c('omni','herbi')) %>% 
  t.test(sleep_rem ~ vore, data=., var.equal=FALSE)
```
## Paired T test
```{r t paired, eval=FALSE, echo=TRUE}
library(PairedData)
# matched exposed and control children
# exposed have parents who worked in factory with lead
head(BloodLead)

t.test(x=BloodLead$Exposed, y= BloodLead$Control, paired=TRUE,
       alternative = "two.sided")
```

## Paired Wilcoxon test
```{r wilcox paired, eval=FALSE, echo=TRUE}
library(PairedData)
# matched exposed and control children
# exposed have parents who worked in factory with lead
head(BloodLead)

wilcox.test(x=BloodLead$Exposed, y= BloodLead$Control,
              paired=TRUE,alternative = "two.sided")
```
```
- Try it out, see if you can interpret results


- p is highly significant
  - can reject the null, accept alternative
  - sample mean 10.43, CI 9.46-11.41





