---
title: "Dplyr Verbs Tutorial"
output: learnr::tutorial
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr)
knitr::opts_chunk$set(echo = FALSE)
library(tidyverse)
library(fivethirtyeight)
data("candy_rankings")
data("starwars")
glimpse(starwars)
glimpse(candy_rankings)
```


## Take a look at Candy data. 
Edit the code below to
Glimpse candy_rankings

```{r candy head, exercise = TRUE}
glimpse(bechdel)
```


## Topic 1 : SeleCting Columns 
Note the C in seleCt - that is how you remember that it selects columns.

*Here's a simple exercise with select*

Edit the R code below to select only the columns:
competitor name, chocolate, and caramel:

```{r select, exercise=TRUE}
candy_rankings %>% 
  select(peanutyalmondy)
```

### More selecting!
The select function has a number of helpers.
You can use the colon to select from varx:vary, including all the columns in between.
YOu can use select(starts_with("c")), or select(ends_with('fraction')).
You can use select(contains("rd"))
You can use select(one_of(c("chocolate", "caramel"))), or
matches(regular expressions), or
num_range(), or everything()

*Here's an exercise with some prepopulated code as well as `exercise.lines = 5` to provide a bit more initial room to work.*

Now select in the code block below:
4 different challenges

1. competitor_name, chocolate, crispedricewafer, and caramel using starts_with
2. the variables that end in percent
3. caramel, hard, bar, and sugarpercent ("ar")
4. variables in a one_of('varname') vector

```{r select2, exercise=TRUE, exercise.lines = 5}
candy_rankings %>% 
  select(competitorname, chocolate, fruity, caramel)
```

## Topic 2 : FilteRing

### Filter
the filteR command has an R in it so that you can remember that you are filtering for Rows.

* Filter for certain rows*

Modify the following code to filter for:

1. candy with a sugar percent > 0.75

```{r filter, exercise=TRUE, exercise.eval=TRUE}
candy_rankings %>% 
  filter(peanutyalmondy == TRUE)
```

### More filtering

Now filter for

1. candy with chocolate in it (note double equals ==)
2. candy with either caramel or nougat (or with | pipe)
3. a bar with a pricepercent >= 0.8
4. a crispedricewafer with a winpercent >0.4
5. a fruity candy with a sugarpercent > 0.7
6. A candy with neither chocolate, caramel, or a fruity component

```{r filtermore, exercise = TRUE}
candy_rankings %>% 
  filter(chocolate == FALSE & caramel == FALSE)
```

## Topic 3 : Mutating

### Mutate
the Mutate command lets you make new variables. For example
mutate(bmi = wt_kg/ht_m^2)

You can do several mutate operations in one set of parentheses, 
separated by commas.

mutate(bmi = wt_kg/ht_m^2,
        risk_score = bmi * total_chol)
        
When you create a new variable with mutate, you can use it immediately in the next statement or in the same data pipe.

Run the code below to see what it does, then
Modify the code to mutate to create:
1. a fruity_caramel variable for candies that have both fruity and caramel ==TRUE

```{r mutate, exercise=TRUE, exercise.eval=TRUE}
candy_rankings %>% 
  mutate(nutty_nougat = round((peanutyalmondy + nougat)/2)) %>% 
  select(peanutyalmondy, nougat, nutty_nougat) %>% 
  arrange(desc(nutty_nougat))
```
 
 If you only want to keep the new variables, use the transmute verb.
```{r transmute, exercise=TRUE, exercise.eval=TRUE}
candy_rankings %>% 
  transmute(nutty_nougat = round((peanutyalmondy + nougat)/2)) %>% 
  arrange(desc(nutty_nougat))
```
### More mutating

Now modify the mutate code below to create

1. sugar_value - the ratio of sugarpercent to pricepercent
2. caramel_bar -candy with both caramel and bar
3. log_winpercent - normalized winpercent by taking natural log
4. sqrt_sugarpercent - normalize this left-tailed variable
5. sugar_pct - convert from a decimal to a percentage.

```{r mutatemore, exercise=TRUE, exercise.eval=TRUE}
candy_rankings %>% 
  mutate(nuttinessperdollar = peanutyalmondy/pricepercent)
```

## Topic 4 : Arranging (Sorting)

### Arrange
the arrange command lets you sort your dataset by particular variables.

You can arrange by several variables in order.
arrange(caramel, chocolate, sugarpercent)

The default is ascending order, but you can use desc() to change that.
arrange(desc(pricepercent))

Run the code below to see what it does, then
Modify the code to arrange the data by:

1. descending winpercent

Note that a select command is added to make it easier to see the result. You can also see this in the whole dataset with View()

```{r arrange, exercise=TRUE, exercise.eval=TRUE}
candy_rankings %>% 
  arrange(sugarpercent) %>% 
  select(competitorname, sugarpercent)
```


### More arranging

Now modify the arrange code below to sort by

1. pricepercent
2. caramel, nougat, desc(sugarpercent)
3. chocolate, descending winpercent 
4. nougat, peanutyalmondy, pricepercent
5. hard, desc(bar), pluribus

```{r arrangemore, exercise=TRUE, exercise.eval=TRUE}
candy_rankings %>% 
  arrange(fruity, desc(sugarpercent))
```

## Topic 5 : Grouping with group_by, then summarize

### group_by
the group_by command lets you group or subset your data by particular variables.
This allows you to easily calculate subgroup summaries or statistics.

You can group by several categorical variables in order.
group_by(caramel, chocolate)

Note that you can't group by continuous variables.

Run the code below to see what it does, then
Modify the code to group the data by chocolate, then calculate the mean winpercent:

Note that a select command is added to make it easier to see the result. You can also see this in the whole dataset with View()

```{r group, exercise=TRUE, exercise.eval=TRUE}
candy_rankings %>% 
  group_by(peanutyalmondy) %>% 
  summarize(mean_win = mean(winpercent))
```


### More grouping

Now modify the arrange code below to 

1. group_by fruity, summarize median pricepercent
2. group_by caramel, summarize max winpercent
3. group_by caramel, then chocolate, summarize mean winpercent
4. group_by bar, then nougat, summarize mean sugarpercent
5. group_by hard, then pluribus, summarize mean winpercent

```{r groupmore, exercise=TRUE, exercise.eval=TRUE}
candy_rankings %>% 
  group_by(bar) %>% 
  summarize(mean_sugar = mean(sugarpercent))
```

## Topic 6 : Tallying, Counting and recounting

### tally
Tally will count the number of observations 
<br>
candy_rankings %>% tally() gives 85
<br>
count will do group_by(var), then tally()
<br>
candy_rankings %>% count(chocolate) counts the number in each subgroup

### count 
the count() command lets you count the number of observations in a particular level of a variable. This is helpful for quickly exploring what you have in each variable, and if there are any strange outliers. You can set sort=TRUE to get a descending sorted table of levels
<br>
count() without a variable argument is the same as tally()
<br>
You can also count() after a count() function to count the number of levels (which can be helpful if you have many of them)

You may want to glimpse the starwars dataset, then
Run the code below to see what it does, then
Modify the code to count by levels of homeworld.

You can also add a denominator variable to your dataset with add_tally() or a count per level by add_count(chocolate)

```{r count, exercise=TRUE, exercise.eval=TRUE}
glimpse(starwars)
starwars %>% 
  count(eye_color, sort = TRUE)
```


### More counting

Now modify the counting code below to 

1. count the number of each species
2. tally the number of observations, and add to dataset with add_tally
3. count the number by eye_color, then add_count to the dataset.
4. count by skin_color and sort
5. count the number by homeworld, and sort

```{r countmore, exercise=TRUE, exercise.eval=TRUE}
starwars %>% 
  count(birth_year, sort=TRUE) 
```

### Quiz

*You can include any number of single or multiple choice questions as a quiz. Use the `question` function to define a question and the `quiz` function for grouping multiple questions together.*

Some questions to verify that you understand the purposes of various base and recommended R packages:

```{r quiz}
quiz(
  question("What symbol do you use to test equality in a logical statement?",
    answer(" !! "),
    answer(" $ "),
    answer(" == ", correct = TRUE),
    answer(" = ")
  ),
  question("What symbol do you use to combine two logical statements with AND?",
    answer(" | "),
    answer(" ^ "),
    answer(" % "),
    answer(" & ", correct = TRUE)
  )
)
```

