---
title: "more_dplyr"
author: "Peter Higgins"
date: "8/15/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(usethis)
library(here)
library(janitor)
```

## More dplyr

You have learned the big 7 verbs in dplyr, including
<br>
1. filteR
2. seleCt
3. mutate
4. group_by
5. summarize
6. arrange
7. count
<br>
But there is more!!

### rename

when you don't need a full mutate, just better names.
This often comes up after you have used clean_names to improve the formatting of variable names, but you end up with long, long names.
<br>
First run the code chunk below to see what you have in the data frame

Then pipe tbl into a rename function, which has the format<br>
rename(new_name = old_name)
and rename the sbp as sbp_basel
```{r rename}
name <- c("Bob", "Carla", "Dave")
the_value_of_sbp_at_baseline <- c(120, 134, 96)
tbl <- data.frame(name, the_value_of_sbp_at_baseline)
tbl %>% 
  rename(sbp_base = the_value_of_sbp_at_baseline)
```

### Re-arrange your variables/columns

If you don't like the order of your variables, you can use *select* to reorder them.
use the format:
select(var1, var2, var3:var4, everything())

note that everything() puts all the other variables in at the end, in their previous order.
This is helpful for pulling interesting variables to the front of the line.

Run the chunk below,
then pipe the tbl into 
 select(pt_id, everything())

```{r rearrange}
firstname <- c("Bob", "Carla", "Dave", "Elena")
pt_id <- c(001, 002, 003, 004)
lastname <- c("Edwards", "Frankel", "Genghis", "Harrison")
tbl <- data.frame(firstname, pt_id, lastname)
tbl %>% 
  select(pt_id, everything())
```

### Find distinct rows

If you want to list only the distinct, unique observations, you can use the distinct function.

Run the code chunk below to see the replicates by visit_num

Then pipe tbl into 
  distinct(var1, var2) - the variables you care about
  
  In this case, use firstname, lastname to find the distinct patients

```{r distinct}
firstname <- rep(c("Bob", "Carla", "Dave", "Elena"),3)
pt_id <- rep(c(001, 002, 003, 004),3)
lastname <- rep(c("Edwards", "Frankel", "Genghis", "Harrison"),3)
visit_num <- c(rep(1,4), rep(2, 4), rep(3,4))
tbl <- data.frame(pt_id, firstname, lastname, visit_num)
tbl %>% 
  distinct(firstname, lastname) ->
unique_pts

```

### Select a group of rows with slice()
Sometimes you just want a few contiguous rows, like with head(), or tail().
but you can also pick which rows with slice()

This is a simple command with 

slice(start_row:end_row) to select the rows you want

Run the chunk below to find out how many rows are in the starwars dataset.

Then look at head and tail functions

Then pipe starwars into a slice function.

Slice out rows 15:25
or
46:60

```{r}
dim(starwars)

starwars %>% 
  head(7)

starwars %>% 
  tail(10)

starwars %>% 
  slice(15:35)
```

### Randomly sample some rows with sample_n() or sample_frac()
Sometimes you want a smaller but representative sample of rows, 
which you can get with sample_n() or sample_frac()

sample_n needs a size (number of rows)

while sample_frac needs a size between 0 and 1.

You can also choose to sample with replacement (replace = TRUE)

Using the chunk below,
try to 

1. sample 30 records from starwars (without replacement)
2. take a 20% random sample (with replacement)

```{r}
starwars %>% 
  sample_n(size =30, replace =TRUE)

starwars %>% 
  sample_frac(size = 0.2, replace = TRUE)
```

but you can also pick which rows with slice()
