---
title: "LearnR blockrand Tutorial"
output: learnr::tutorial
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr)
knitr::opts_chunk$set(echo = FALSE)
if (!require(blockrand)){    # checks if already installed
        install.packages(blockrand) #installs if not
}
library(blockrand)
library(knitr)
library(kableExtra)
library(tidyverse, quietly = T)
```


## Topic 1

### Exercise 

*Here's a simple exercise with an empty code chunk provided for entering the answer.*

Write the R code required to run a block randomization with n=200:

```{r ex1, exercise=TRUE}

```

### Exercise with Code

*Here's an exercise with some prepopulated code*
Now edit the R code below to randomize 120 subjects to 
upadacitinib, risankizumab, or upadacitinib + risankizumab:

```{r ex2, exercise=TRUE}
blockrand(n = 100,
         block.prefix = "B_", # this is for block.id
         stratum = "Stratum_1",
         num.levels =  , # N distinct treatments
         levels = c("A", "B", "C")
          ) 
```

## Topic 2

### Exercise with Hint

*Here's an exercise where the chunk is pre-evaulated via the `exercise.eval` option (so the user can see the default output we'd like them to customize). We also add a "hint" to the correct solution via the chunk immediate below labeled `print-limit-hint`.*

Modify the following code to randomize 250 patients to 4 doses of upadacitinib (7.5, 15, 30, 45 mg) vs. placebo:

```{r ex3, exercise=TRUE, exercise.eval=TRUE}
blockrand(n = 100,
         block.prefix = "B_", # this is for block.id
         num.levels =  17, # N distinct treatments
         levels = c("A", "B", "C")
          ) 
```

```{r ex3-hint}
num.levels = 5
```

### Quiz

*Use help or google to find the documentation for blockrand. Figure out how each argument works, and what each is for.*

Some questions to verify that you understand the purposes of different arguments to blockrand:

```{r quiz}
quiz(
  question("Which argument sets the number of arms in the clinical trial?",
    answer("n"),
    answer("block.prefix"),
    answer("levels"),
    answer("num.levels", correct = TRUE)
  ),
  question("Which argument names/labels the arms in the clinical trial?",
    answer("n"),
    answer("block.prefix"),
    answer("levels", correct = TRUE),
    answer("num.levels")
  )
)
```

