---
title: "Categorical vs Continuous Boxplots and More"
author: "Peter Higgins"
date: "6/14/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(pacman)
p_load(tidyverse, fivethirtyeight, HSAUR, ggsignif, ggridges, scales, ggplot2)
```

## Boxplots
Boxplots are common ways to compare a continous variable across several categories, and are often used to illustrate the continous outcome of an experiment across several groups.
<br>
The basic geom is geom_boxplot, with the aes containing a categorical variable (usually x),
and a continuous variable, usually y. As usual, geom_boxplot inherits aes from the ggplot function.
<br>
Arguments to geom_boxplot include
- outlier.color, outlier.fill
- outlier.shape, outlier.stroke, outlier.alpha
- outlier.size
- color
- fill
- position = 'dodge2'

<br>
We will use the data from 4 clinical trials published from 1987-1989 by Lanza on the use of misoprostol to reduce the severity of NSAID-induced gastric ulcers on a 1-5 scale.
Start by taking a look at the data.
<br>
Details are here: https://vincentarelbundock.github.io/Rdatasets/doc/HSAUR/Lanza.html 

```{r mpg}
data(Lanza)
glimpse(Lanza)
```

## First Boxplot
 Your basic boxplot aes needs a categorical x variable, and a continuous y variable. 
  Add a geom_boxplot() to the code below to get the basic boxplot by treatment group.
  Then use the color and fill arguments to map treatment to these graphic aesthetics. Use the show.legend=FALSE argument to geom_boxplot to control the legend. 

```{r boxplot1, echo=FALSE}
Lanza %>% 
  mutate(ulcers = as.numeric(classification))%>%  ggplot( aes(x = treatment, y = ulcers) ) 
```

## Second Boxplot, with grouping by trial
You can map another variable to your aesthetic statement with the fill aes. In this case, we will use x= study and fill= treatment, to separate the treatments the data in each study. Edit the code below to add the aes(fill=) mapping to produce comparisons for all 4 studies by treatment group. 

```{r boxplot2, echo=FALSE}
Lanza %>% 
  mutate(ulcers = as.numeric(classification))%>%  ggplot(aes(x = study, y = ulcers)) +
  geom_boxplot()
```

## Third Boxplot, with grouping by trial using facets
Another way to do subgrouping of your data is to facet it into small multiples. This produces multiple small graphs. You can facet by a single variable with facet_wrap(~newvar) In this case, we will use x = treatment, y = ulcers, and add facet_wrap(~ study), to separate the data in each study. Edit the code below to produce comparisons for all 4 studies by treatment group.  Experiment with which variable is x and which is the variable used for faceting.

```{r boxplot3, echo=FALSE}
Lanza %>% 
  mutate(ulcers = as.numeric(classification))%>%  ggplot(aes(x = treatment, y = ulcers)) +
  geom_boxplot() +
  facet_wrap(~study)
```

<br>
<br>
<br>
<br>
## CHALLENGE 1: make your own boxplot with a new dataset
Without looking at or copy/pasting from the previous code,
use the data from the polyps dataset to make a boxplot comparing the number of polyps in FAP patients taking placebo vs. taking NSAIDs.
<br>
Run the code below, then edit it to produce the desired boxplot.
Remember to:

1. map the aesthetics to both a categorical (treat) and a continuous (number of polyps) variable
2. add the right geom
3. Add title, labels with labs()
4. use constant color and fill, or map these to variables also.
5. Style the theme, colors, base_size as desired.


```{r}
data("polyps", package = "HSAUR")
glimpse(polyps)

ggplot(data=polyps)
  
```




<br>
<br>
<br>
<br>

## Fourth Boxplot, with grouping by trial and treatment using facets
You can facet by two variables by adding facet_grid(catvar1 ~ catvar2). Try faceting by categorical variables treatment and study. Try again in reverse order. 

```{r boxplot4, echo=FALSE}
Lanza %>% 
  mutate(ulcers = as.numeric(classification))%>%  ggplot(aes(y = ulcers)) +
  geom_boxplot() 
```

## Fifth Boxplot, using arguments for varwidth, color, fill, and notch


```{r boxplot5, echo=FALSE}
Lanza %>% 
  mutate(ulcers = as.numeric(classification))%>%  ggplot(aes(x = treatment, y = ulcers, fill = study)) +
  geom_boxplot( varwidth = TRUE, notch = TRUE,
        color = 'darkorchid1', fill = 'white') 
```

## Sixth Boxplot, with overlay of datapoints with geom_jitter and remove outliers (no need to repeat these). This provides a clearer view of the data.
What would happen if you put geom_boxplot last?
What would happen if outlier.shape = NA was removed?

```{r boxplot6, echo=FALSE}
Lanza %>% 
  mutate(ulcers = as.numeric(classification))%>%  ggplot(aes(x = treatment, y = ulcers)) +
  geom_boxplot( varwidth = TRUE,
        color = 'steelblue', fill = 'lightblue',
        alpha = 0.3, outlier.shape = NA) +
  geom_jitter(width = 0.3, alpha = 0.6) +
  theme_bw(base_size = 16)
```

## Seventh plot, with violin
You can see how a boxplot obscures the distribution. Try this again with a violinplot. change the boxplot geom to violin in the code below.

```{r boxplot7, echo=FALSE}
Lanza %>% 
  mutate(ulcers = as.numeric(classification))%>%  ggplot(aes(x = treatment, y = ulcers)) +
  geom_boxplot(color = 'magenta', 
              fill = 'lightblue',
        alpha = 0.7) +
  geom_jitter(width = 0.2, alpha = 0.6) +
  theme_bw(base_size = 16)
```

<br>
<br>
<br>
<br>
## CHALLENGE 2: make your own violinplot overlaid with jittered datapoints with a new dataset
Without looking at or copy/pasting from the previous code,
use the data from the polyps dataset to make a violin plot comparing the number of polyps in FAP patients taking placebo vs. taking NSAIDs.
<br>
Run the code below, then edit it to produce the desired boxplot.
Remember to:

1. map the aesthetics to both a categorical (treat) and a continuous (number of polyps) variable
2. add the violin geom
3. Overlay with the jitter geom, control the width
4. Add title, labels with labs()
5. use constant color and fill, or map these to variables also.
6. Use alpha to adjust the intensity of the violin fill and of the jittered points.
7. Style the theme, colors, base_size as desired.

```{r}
data("polyps", package = "HSAUR")
glimpse(polyps)

ggplot(data=polyps)
  
```

<br>
<br>
<br>
<br>
## Eighth Plot, with ridgeplots
 Ridgeplots are a variant of violin plots that look like ridges, or mountain ranges. These are nice for comparing distribution densities.
 You need to map your continuous variable in the x aesthetic, then pick categorical variables for your y and /or your fill aesthetics.
 Map ulcers to x, study to y, and treatment to fill.  Experiment with swapping y and fill mappings.
 

```{r plot8, echo=FALSE}
Lanza %>% 
  mutate(ulcers = as.numeric(classification)) ->
lanza2
lanza2 %>% 
ggplot(aes(x = ulcers, y = study, fill = treatment)) +
  geom_density_ridges()
```

## Ninth plot, with ridges in series
Ridgeplots are often use to illustrate changes in distributions over time (consecutive months, or years).
Add facet_wrap(~treatment) and see the distribution change by treatment from study to study.
Add ncol=1 to the facet_wrap for a different view

```{r plot9, echo=FALSE}
lanza2 %>% 
  ggplot(aes(x = ulcers, y = study)) +
  geom_density_ridges( fill = 'magenta') 
```

## Tenth plot, with raincloud plots
You can add the points (the 'rain') to the clouds by adding the argument jittered_points = TRUE to geom_density_ridges.
You can also have the rain 'fall' by adding the arguments jittered_points = TRUE, position='raincloud', alpha =0.5, scale =0.9 <br>
to geom_density_ridges.

```{r plot10, echo=FALSE}
lanza2 %>% 
ggplot(aes(x = ulcers, y = study, fill = treatment)) +
  geom_density_ridges() +
  theme(legend.position = 'top')
```
<br>
<br>
<br>
<br>
## CHALLENGE 3: make your own raincloud plot with a new dataset
Without looking at or copy/pasting from the previous code,
use the data from the polyps dataset to make a raincloud plot comparing the age in FAP patients taking placebo vs. taking NSAIDs.
<br>
Run the code below, then edit it to produce the desired boxplot.
Remember to:

1. map the aesthetics to both a categorical (treat) and a continuous (age) variable
2. add the raincloud geom
3. Use jittered_points = TRUE, position='raincloud'
4. adjust the alpha and scale options
5. use constant color and fill, or map these to variables also.
6. Use alpha to adjust the intensity of the jittered points.
7. Style the theme, colors, base_size as desired.
8. add appropriate labels

```{r}
data("polyps", package = "HSAUR")
glimpse(polyps)

ggplot(data=polyps)
  
```

<br>
<br>
<br>
<br>

## Eleventh plot, with mean labels
You can label the mean in a boxplot with a stat_summary, using function fun.y = mean, and specifying the geom as text. It helps to round the y to 1 decimal place.
<br>
more details here https://stackoverflow.com/questions/19876505/boxplot-show-the-value-of-mean 

```{r plot11, echo=FALSE}
lanza2 %>% 
  ggplot(aes(x = treatment, y = ulcers)) +
    geom_boxplot(color = 'darkorchid1',         fill = 'white') +
  stat_summary(fun.y = mean, geom = 'text', aes(label = round(..y.., digits =1))) +
  theme_classic(base_size = 15)
  
```

## Twelfth plot, with ggsignif
You can make comparisons and list p values between groups with the ggsignif package/
<br>
more details here: 
https://cran.r-project.org/web/packages/ggsignif/vignettes/intro.html


```{r plot12, echo=FALSE}
lanza2 %>% 
  ggplot(aes(x = treatment, y = ulcers)) +
    geom_violin(color = 'steelblue',         fill = 'lightblue', alpha =0.7) +
  geom_boxplot(width =0.2, alpha =0.3) +
  geom_jitter(width = 0.2, alpha =0.2) +
  theme_classic(base_size = 15) +
  geom_signif(comparisons = list(c("Misoprostol", "Placebo")),
              y_position = 5.8) +
  ylim(0.4,6.3)
```

<br>
<br>
<br>
<br>
## CHALLENGE 4: make your own boxplot with significance comparison and p value for a new dataset
Without looking at or copy/pasting from the previous code,
use the data from the polyps dataset to make a boxplot/violinplot comparing the number of polyps in FAP patients taking placebo vs. taking NSAIDs.
<br>
Run the code below, then edit it to produce the desired plot.
Remember to:

1. map the aesthetics to both a categorical (treat) and a continuous (polyps) variable
2. add the violin geom first, then boxplot
3. adjust the alpha and width options for violin
4. add ggsignif, list the correct groups for comparison, set the y position correctly 
5. use constant color and fill, or map these to variables also.
6. Style the theme, colors, base_size as desired.
7. adjust ylim if needed
8. add appropriate labels

```{r}
data("polyps", package = "HSAUR")
glimpse(polyps)

ggplot(data=polyps)
  
```
