---
title: "Introducing Strings"
output: learnr::tutorial
runtime: shiny_prerendered
---

Let's learn about strings!

## Data Types
Strings (or character class) are one of the basic data types (classes) in R.

Other data types include:

1. Logical (TRUE/FALSE)
2. integer (0/1/2/3/4)
3. numeric (3.2, -7.14, 11.78)
4. factor (red/blue/green) with variants:
  a. ordinal factor (low/medium/high)
  b. factor with labels (1/2, labels = c("female", "male"))
5. complex (3 + 2i, 7 - 3i)
6. (and of course) strings ("Bob", "yellow", "3.14")

anything can be a string - so it is often a default when reading in data of **mixed** types. 

Let's get started - run the setup chunk below to set up your libraries.
Execute this chunk by clicking the green triangle *Run* button (top right) within the chunk.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(tidyverse)
```

## A first string vector
Here is an example of a vector of six strings.
Run the code with the green triangle to see the `strings` object, its length, and its underlying class.

```{r strings-1, exercise = TRUE}
strings <- c(c("alcoholic cirrhosis", "cirrhotic", "alcoholism", "alcohol abuse disorder", "alcohol toxicity", "alcoholic hepatitis"))
strings
length(strings)
class(strings)
```

This gives you three sets of results

1. the vector of 6 strings
2. the length of the strings object (6)
3. the class of the strings object (character)
Note that each of these results are vectors (even if only of length 1), and are numbered.

## The Good and the Bad of Coercion
Variables in a data frame in R are all vectors.
Each column represents a vector.

In a vector, all data elements have to be the same data type.
If they are not, they will be "coerced" to a common data type.
This can be convenient at times, but can also be **dangerous** if you don't see it coming. 
Below are some illustrative examples of mixed data and coercion in a code chunk.

Predict what you think the classes of the vectors `values` and `numbers` will be in the chunk below. Remove the `#` from the final line, `numbers * 2`, then run each of the 3 groups of lines in the code chunk below by selecting the group, then pressing *Cmd+Enter*. Check your predictions.

```{r strings-2, exercise = TRUE} 
values <- c(TRUE, FALSE, 7)
numbers <- c(3, 9, " ")
values
class(values)

numbers
class(numbers)

pi * values^2 
#numbers * 2
```
Note that you can multiply TRUE and FALSE values (in `values` because these will be coerced to 1 and 0 when in a vector with a numeric), and an extra space in a column (often seen in data imported from Excel) can cause coercion of a numeric variable to the character class (as seen in `numbers`), making attempts to multiply or otherwise numerically manipulate this variable throw an error like `Error in numbers * 2 : non-numeric argument to binary operator`.

## Stringr is your multitool for working with Strings

Now that you have seen this common _'gotcha'_ with strings, on to things you can do with strings! 

The tidyverse _stringr_ package has a lot of helpful string functions.

Try help(stringr) to see some of them

Note that you can get a lot more detail (and a cheatsheet)
at (click on the link)

[The Stringr Website](http://stringr.tidyverse.org)

Or within RStudio (check this out in the menus above) at Help/Cheatsheets/Browse cheatsheets...and find the `stringr` cheatsheet.

## Finding Patterns in Strings
### str_detect()
Let's start with the strings vector from above
use _str_detect()_ to find "alc" and "cirr" in the vector
Run the code block below for an example

```{r str-detect, exercise = TRUE}
strings
str_detect(strings, "alc")
str_detect(strings, "cirr")
```
This is somehat helpful to get a logical vector of `TRUE` and `FALSE` back (this will turn out to be more helpful later).

### str_which()
Now try the _str_which()_ function - add code to the block below to find which data elements contain "alc" and "cirr"

The syntax is helpfully consistent _str_which_(`string_to_be_searched`)

```{r str_which, exercise = TRUE}
strings


```

### str_count()
Now try _str_count()_ - add code to the block below to find how many times these data elements contain "alc" and "c"

```{r str-count, exercise = TRUE}
strings


```

### str_locate()
Now try _str_locate()_ - add code to the block below to find where these data elements contain "alc", "cirr", and "c"

```{r str-locate, exercise = TRUE}
strings



```

## Subsetting Text
### str_sub()
Now use _str_sub()_ to subset bits of the text.
The arguments of str_sub() are 
str_sub(`text`, `start_integer`, `end_integer`)
Try both positive and negative (small) integers

```{r str-sub, exercise = TRUE}
strings



```

### str_subset()
_str_subset()_ does something different - it gives you a subset of the vector, based on matches.
Tray str_subset with "alc" and "cirr"
```{r str-subset, exercise = TRUE}
strings



```

### str_extract()
_str_extract()_ does something different - it gives you a subset of the data elements, based on matches.
Try str_extract with "alc" and "cirr"
str_match is pretty similar - give that a try as well
Can you tell the difference?

```{r str-extract, exercise = TRUE}
strings

```
### str_match_all()
also try _str_match_all()_ with "alc" as your pattern

```{r str-match-all, exercise = TRUE}
strings

```

## Measuring and Changing String Length
### str_length()
Sometimes you want to measure or control the length of strings
try _str_length()_ with strings,
and with mrns (medical record numbers)
```{r str-length, exercise = TRUE}
mrns <- c("012345678", "23456789     ", "345678901   ", "123456")


```

### str_trunc()
when the strings are too long, you can truncate them
with _str_trunc_(string, length)
Try this with strings and mrns below
```{r str-trunc, exercise = TRUE}

```

### str_pad()
Sometimes leading zeros get left off of numbers, especially in Excel or with MRNs.
You can pad these to the right length with
_str_pad_(text, length)
This defaults to padding with spaces on the left.
You can add a side argument for "right", "both", or "left" to str_pad
i.e. side = "left"
and a pad argument for what character to pad with
i.e. pad = "0"
Try to get all the MRNs to length 9,
with the right padding character on the correct side

```{r str-pad, exercise = TRUE}

```

### str_trim()
If you just want to trim off extra whitespaces,
you can use _str_trim(string)_
Try this with mrns
```{r str-trim, exercise = TRUE}
mrns
str_trim(mrns)

```

## Changing Srings
### str_replace()
You can also use _str_replace()_ 
to replace some text in string elements
The three arguments are _string_, _pattern_, and _replacement_.
Try the replacement below.
Dream up some other options and try them out
```{r str-replace, exercise = TRUE}
str_replace(strings, "cirr", "whirr")
```

### str_remove()
You can also use _str_remove()_ 
to remove some text pattern from string elements.
The two arguments are _string_ and _pattern_.
Try removing the pattern "alc" from strings in the chunk below.
Dream up some other options and try them out
```{r str-remove, exercise = TRUE}
strings
```

### str_wrap()
You can also change strings with _str_wrap()_ 
to srap strings into nice paragraphs. 
The four arguments are _string_, _width_, _indent_, and _exdent_.
Try wrapping the strings6 into a nice looking paragraph with indents and an exdent in the code chunk below.

```{r str-wrap, exercise = TRUE}
strings6 <- str_c(rep(strings, 6), collapse = ", ")
strings6 
```

## Changing the Case of sTrINGs
### str_to_title(), str_to_upper(), str_to_lower()
If you have text that has an unruly case,
like ALCOHolic CiRRhosIS
you can fix that with a few commands <br>
_str_to_title()_, <br>
_str_to_upper()_, <br>
_str_to_lower()_, <br>
_str_to_sentence()_, <br>
all help
Try changing strings to Upper case, title case, lowercase, then Sentence case in the code block below.
```{r case, exercise = TRUE}
strings

```

## Combining and Splitting Strings

### Combining Strings with str_c()
Sometimes you get strings as a vector of string chunks, but you actually want them as a single character string.
This can be fixed with _str_c()_ for combine (or collapse)
Try the chunk below, then change the collapse character to 
three spaces, or space-underscore-space, or just a single space.

```{r str-c, exercise = TRUE}
str_c(strings, collapse = "+")
```


### Splitting Strings with str_split()
Sometimes you get strings that combine two data fields, but you actually want them separated into different variables. Common examples include Name (combines firstname and lastname) and Month-Day, which contains two distinct variables.
This can be fixed with _str_split()_.
_str_split()_ has two main arguments, _string_ and _pattern_.
Review the small `bdays` dataframe in the chunk below, then split the two variables using str_split() on the patterns space " " or dash "-" by running the code chunk.


```{r str-split, exercise = TRUE}
bdays <- data.frame(Name = c("Bob Anderson", "Cindy Davis", "Eric Fields", "Gemma Harris"),
           Birthday = c("Jan-22", "May-14", "Apr-04", "Dec-11"))
bdays$Name
bdays$Birthday  %>% str_split("-")
```

### Splitting strings to separate variables with separate()
The _separate()_ function is in the **tidyr** package, because it is so often needed in tidying datasets. It conveniently splits variables that combine data into distinct new variables.
_separate()_ has 4 main arguments: data, col (column to split), into (names of new resulting columns), and sep (which defines the separator to split on).
Run the example below for the Name column, then write code for a similar `separate` function on the Birthday column in the code chunk below.
Try to end up with 4 distinct columns for the bdays dataset.

```{r separate, exercise = TRUE}
bdays %>% separate(col = Name, into = c("firstname", "lastname"), sep=" ")
```

```{r separate-hint-1}
bdays %>% 
  separate(col = Name, into = c("firstname", "lastname"), sep=" ") %>% 
  separate(col = Birthday, into = c("month", "day"), sep="-") 
```

