---
title: "Randomization with blockrand in R"
author: "Peter Higgins"
date: "4/22/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(blockrand)
library(tidyverse)
```

# Randomization for Clinical Trials with R

There are a number of packages for doing key functions for clinical trials in R. You can find many of these on the CRAN Task View for Clinical Trials, at <https://cran.r-project.org/web/views/ClinicalTrials.html>.

This is a curated list of packages that anyone might find useful in designing, monitoring, or analyzing clinical trials, and is often a good place to start in looking for packages that might be relevant for clinical trials.

If you use `Ctrl-F` to search the web page for "rand", several packages address randomization, including

-   blockrand

-   randomizeR

-   pwr

-   experiment

-   clusterPower

-   CRTSize

-   cosa

-   PowerupR

Several of these are specifically for more complex designs, including cluster and multilevel randomization (clusterPower, cosa, CRTSize). For today, we will focus on the straightforward randomization packages including {blockrand} and {randomizer}. The {blockrand} package creates randomizations for clinical trials with can include stratified enrollment and permuted block randomization, and can produce a PDF file of randomization cards.

Let's start with an example in {blockrand}.

Details on the package can be found at <https://cran.r-project.org/web/packages/blockrand/blockrand.pdf>

or by running `help(blockrand)` in your Console.

You want to randomize 180 inpatients with severe ulcerative colitis to one of 3 arms: corticosteroids alone (control), corticosteroids + tofacitinib, or corticosteroids + upadacitinib. You want to stratify the participants by (1) prior biologic failure and (2) Albumin level above or below 3.0. To be prepared for dropouts and imbalanced enrollment, you want to have a randomization list with at least 60 assignments available for each arm and stratum. To avoid a recognizable pattern in the randomization, you want to have a permuted block design with blocks of sizes 3, 6, and 9.\
Below, you will see how to do this for the biologic failure - low albumin stratum.

```{r uc-study-1}
bfla <- blockrand(n = 60, 
                     num.levels = 3, # three treatments
                     levels = c("CS", "CS/Tofa", "CS/Upa"), # arm names
                     stratum = "Bfail.LowAlb", # stratum name
                     id.prefix = "BfLA", # stratum abbrev
                     block.sizes = c(1,2,3), # times arms = 3,6,9
                     block.prefix = "BfLA") # stratum abbrev
bfla
```

You can see the `id` for each participant, their `stratum`, the `block.id` for their permuted block, the `block.size`, and their assigned `treatment`. You can imagine this as a randomization list, or as assignments that you could print out on cards and seal in security envelopes for the time of randomization. Of course, this is only one of our four strata. We should do the same for the 3 other strata.

```{r uc-study-2}
bfha <- blockrand(n = 60, 
                     num.levels = 3, # three treatments
                     levels = c("CS", "CS/Tofa", "CS/Upa"), # arm names
                     stratum = "Bfail.HiAlb", # stratum name
                     id.prefix = "BfHA", # stratum abbrev
                     block.sizes = c(1,2,3), # times arms = 3,6,9
                     block.prefix = "BfHA") # stratum abbrev
bfha

bnha <- blockrand(n = 60, 
                     num.levels = 3, 
                     levels = c("CS", "CS/Tofa", "CS/Upa"),
                     stratum = "Bnaive.HiAlb",
                     id.prefix = "BnHA",
                     block.sizes = c(1,2,3, 4), 
                     block.prefix = "BnHA")
bnha

bnla <- blockrand(n = 60, 
                     num.levels = 3, 
                     levels = c("CS", "CS/Tofa", "CS/Upa"),
                     stratum = "Bnaive.LoAlb",
                     id.prefix = "BnLA",
                     block.sizes = c(1,2,3), 
                     block.prefix = "BnLA")
bnla
```

## Printing these on Cards

Ideally, you will print out each randomization on a card, and seal it in a security envelope, with the outside of the envelope labeled with the id. You can do this with the *plotblockrand()* function. This function creates a pdf file of randomization cards for printing. These are designed so that the middle text will show in a standard letter sized envelope with a window, and the top text (the assignment) can be folded over to increase security (against anyone trying to peek through the security envelope to guess the assignment).

```{r print-cards}
uc_study <- bind_rows(bfha, bfla, bnha, bnla) # bind together the four strata into one dataframe

blockrand::plotblockrand(uc_study, # input dataframe
                    file = "uc_study.pdf", # output pdf
                    # top hidden text with assignment
      top=list(text=c('My Study','Patient: %ID%','Treatment: %TREAT%'),
                col=c('black','black','red'),font=c(1,1,4)),
      # middle text to show through window of # 10 envelope 
    middle=list(text=c("My Study","Stratum: %STRAT%","Patient: %ID%"),
               col=c('black','blue','orange'),font=c(1,2,3)),
    # bottom text- any instructions to study coordinator
   bottom="Call 123-4567 to report patient entry",
   cut.marks=TRUE) # add cut marks - 4 per page

```

Open up the file `uc_study.pdf` in your Files tab to see the output pdf file, with assorted fonts and colors.

Just for fun, change (and then re-run) the

-   text "My Study" to something more interesting

-   change "Patient" to "Participant"

-   change "Treatment" to "Arm" or "Assignment"

-   change some of the colors to standard R colors

-   change some of the fonts (within 1-4)

Sometimes with equal blocks, and clear treatment effects or side effects, nurses or study coordinators can guess the randomization pattern. If you want to get fancy, and make it even harder to guess treatment assignments, you can add one of the unequal blocks options, to make it hard to find patterns in treatment or in side effects. Set `uneq.beg = TRUE` for an unequal block in the beginning, or `uneq.mid = TRUE` for an unequal block in the middle.

## Now, try this yourself

You want to randomize 80 outpatients with Crohn's disease to one of 8 arms, as part of a 2\^3 factorial design to increase patient activation. These arms involve using (A,B, C) or not using (a,b,c) 3 intervention components. The 8 arms then become:

-   abc

-   abC

-   aBc

-   aBC

-   Abc

-   AbC

-   ABc

-   ABC

Then we want to stratify the participants by baseline PAM score (a measure of patient activation) with levels of low, medium, and high PAM.

To be prepared for dropouts and imbalanced enrollment, you want to have a randomization list with at least 32 assignments available for each arm and stratum. To avoid a recognizable pattern in the randomization, you want to have a permuted block design with blocks of sizes 8 and 16.

::: {.tryit}
You can hover over top right corner of the code chunk below, and a copy icon will appear - click this to copy the code to your clipboard.
You can then paste it into your local version of RStudio, edit it, and run it.
:::

In the code block below, fill in the blanks to complete the code to make a dataframe for the `low_pam` stratum.

```{r crohn-study-1, error=TRUE, eval=FALSE}
low_pam <- blockrand(n = __, 
                     num.levels = __, #eight treatments
  levels = c("abc", "abC", "aBc", "aBC",
             "Abc", "AbC", "ABc", "ABC"), # arm names
                     stratum = "__", # stratum name
                     id.prefix = "lp", # stratum abbrev
                     block.sizes = c(1,2,3), # times arms 
                     block.prefix = "LP") # stratum abbrev
low_pam
```

Now that you have one stratum sorted, edit the code block below to create the `med_pam` and `high_pam` strata.

```{r crohn-study-2, error=TRUE, eval=FALSE}
med_pam <- blockrand(n = __, 
                     num.levels = __, #eight treatments
  levels = c("abc", "abC", "aBc", "aBC",
             "Abc", "AbC", "ABc", "ABC"), # arm names
                     stratum = "__", # stratum name
                     id.prefix = "__", # stratum abbrev
                     block.sizes = c(__), # times arms 
                     block.prefix = "__") # stratum abbrev
med_pam

high_pam <- blockrand(n = __, 
                     num.levels = __, #eight treatments
  levels = c("abc", "abC", "aBc", "aBC",
             "Abc", "AbC", "ABc", "ABC"), # arm names
                     stratum = "__", # stratum name
                     id.prefix = "__", # stratum abbrev
                     block.sizes = c(__), # times arms 
                     block.prefix = "__") # stratum abbrev
high_pam
```

Great!

Now try to

-   bind these 3 strata into one dataframe

-   print these as cards to a pdf file

Edit the code chunk below to produce the pdf file

```{r print-crohn-cards, error=TRUE, eval=FALSE}
cd_study <- bind_rows(__,__,__) # bind together the 3 strata into one dataframe

blockrand::plotblockrand(__, # input dataframe
                    file = "cd_study.pdf", # output pdf
                    # top hidden text with assignment
      top=list(text=c('CD Study','Patient: %ID%','Treatment: %__%'),
                col=c('orange','blue','red'),font=c(1,1,4)),
      # middle text to show through window of # 10 envelope 
    middle=list(text=c("CD Study","Stratum: %STRAT%","Patient: %__%"),
               col=c('black','red','cadetblue'),font=c(1,2,3)),
    # bottom text- any instructions to study coordinator
   bottom="Call 123-4567 to report patient entry",
   cut.marks=TRUE) # add cut marks - 4 per page

```

## Now Freestyle

Your turn. Create randomization tables and a pdf file of cards for a study of 2 microbiome interventions to reduce the formation of colon adenomas.

-   your 3 study arms will be - placebo, Streptococcus thermophilus, and S.thermo plus lactose (a preferred sugar for S.t, making this arm a synbiotic, while arm 2 is a probiotic) - aka 3 arms called: pbo, probiotic, synbiotic.

-   Your stratifications will be by

    -   prior polyps being MSI_hi or MSI_lo (for microsatellite instability mutations)

    -   BMI above or below 35. BMI_hi, BMI_low

-   block sizes of 4,8,12,16

-   160 per arm

Edit the code block below for the first stratum

```{r adenoma-study-1, error=TRUE, eval=FALSE}
mhbh <- blockrand(n = __, # treatment arms
                     num.levels = __, # of treatments
  levels = c("placebo", "probiotic", "synbiotic"), # arm names
                     stratum = "__,__", # stratum name
                     id.prefix = "mhbh", # stratum abbrev
                     block.sizes = c(__,__,__,__), # times arms 
                     block.prefix = "__") # stratum abbrev
mhbh
```

Edit the code block below for the remaining strata

```{r adenoma-study-2, error=TRUE, eval=FALSE}
mhbl <- blockrand(n = __, # treatment arms
                     num.levels = 3, # of treatments
  levels = c("placebo", "probiotic", "__"), # arm names
                     stratum = "msi_hi.bmi_lo", # stratum name
                     id.prefix = "__", # stratum abbrev
                     block.sizes = c(1,2,__,__), # times arms 
                     block.prefix = "MHBL") # stratum abbrev
mhbl

mlbl <- blockrand(n = 160, # treatment arms
                     num.levels = __, # of treatments
  levels = c("placebo", "__", "synbiotic"), # arm names
                     stratum = "__", # stratum name
                     id.prefix = "mlbl", # stratum abbrev
                     block.sizes = c(__,__,3,4), # times arms 
                     block.prefix = "MLBL") # stratum abbrev
mlbl

mlbh <- blockrand(n = __, # treatment arms
                     num.levels = 3, # of treatments
  levels = c("__", "probiotic", "synbiotic"), # arm names
                     stratum = "msi_lo.bmi_hi", # stratum name
                     id.prefix = "__", # stratum abbrev
                     block.sizes = c(1,2,3,4), # times arms 
                     block.prefix = "MLBH") # stratum abbrev
mlbh

```

Edit the code block below to bind the strata together and print the cards

```{r adenoma-study-cards, error=TRUE, eval=FALSE}
adenoma_study <- bind_rows(mlbl, mlbh, mhbh, mhbl) # bind together the strata into one dataframe

blockrand::plotblockrand(__, # input dataframe
                    file = "adenoma_cards.pdf", # output pdf
                    # top hidden text with assignment
top=list(text=c('Adenoma Study','Patient: %__%','Treatment: %TREAT%'),
                col=c('orange','blue','red'),font=c(1,1,4)),
      # middle text to show through window of # 10 envelope 
    middle=list(text=c("Adenoma Study","Stratum: %__%","Patient: %ID%"),
               col=c('black','red','cadetblue'),font=c(1,2,3)),
    # bottom text- any instructions to study coordinator
   bottom="Call 123-4567 to report patient entry. \nInstruct participant to avoid antibiotics and stop aspirin",
   cut.marks=TRUE) # add cut marks - 4 per page


```
