---
title: "Secure Passwords in R"
author: "Peter Higgins"
date: "5/6/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(keyring)
library(rdrop2)
```

# Secure Passwords in R

You will, not infrequently, need to use passwords to access secure or PHI protected data from databases or cloud storage or even password-protected excel files.

It is important to sort out how to use passwords securely. You do not want to type out your password in a script. You can store passwords securely in variables on your computer for use later. If they are in a script, someone else can find them, especially if you are using version control and pushing code to Github.

You *can* use the **.gitignore** file to make sure that git ignores certain files to back up to the repository, especially large files (graphics, etc. that can be regenerated), but it is better to avoid putting any passwords into your code at all.

Each major computer operating system (Windows, MacOS, Linux) has its own secure credential store. You may know this as the 'Credential Store' on Windows, the 'Keychain' on MacOS, or the "Secret Service API' on Linux. Each of these is a secure, password-protected store of passwords that you use for different websites for browser access.

The {keyring} package in R gives you a single interface to all of these secure credential stores. Make sure you have this installed with<br>

`install.packages('keyring')`

after you call `library(keyring)`, you can store one or more secret passwords (keys). A key is defined by a service name (i.e. website) and a password. Once defined, this key persists in the keyring store of your operating system, and is only accessible to someone with the password to your computer.

You can define a key once (interactively, or in an R script), and it will persist for use in future R sessions.

Let's see your current list of secrets. Run the code chunk below.

```{r keyring}
library(keyring)
key_list()
```

This gives you a list of all the keys stored by your operating system credential store.

## Setting New Keys

You can add a new secret with *key_set()*

```{r , eval=FALSE}
key_set(service = "secure_web", username = "pdrh")
```

This will open an interactive window for you to securely enter your hidden password. This is stored in your credential store.

If you now run key_list(), and store it in a new object named keys, you can View this object and find new secret, named secure_web. Run the code chunk below.

```{r , eval=FALSE}
keys <- key_list()
View(keys)
```

You may need to sort these secrets by service (click on the header of this column) to make it easier to find. Note that you see the service and username, but the password is not visible. To read the secret, you need to use the *key_get*() function.

```{r eval = FALSE}
secure_web_pwd <- key_get(service = "secure_web", username = "pdrh")
secure_web_pwd
```

This actually returns the password, in this case, to the console.

You can use this to respond to a password request, or pass the password to a web API. These often use a function to retrieve data that requires the username and password as arguments.

For the imaginary website, secure_web, we can use this approach to access web storage and retrieve files.

```{r eval = FALSE}
secure_web_pwd <- key_get(service = "secure_web", username = "pdrh")
get_secure_web_files(username = "pdrh", password = secure_web_pwd)
```

If you want to delete a secret, you can use key_delete()

```{r eval = FALSE}
key_delete(service = "secure_web", username = "pdrh")
```

You can check this by regenerating your keys dataframe

```{r , eval=FALSE}
keys <- key_list()
View(keys)
```

You will find that the secret for secure_web is now gone from your Credential Store.

You can also use an R package named {credentials}, which is also popular, and it stores your credentials within git.

<https://happygitwithr.com/credential-caching.html>
