---
title: "Mutating Joins to Combine Data Sources"
author: "Peter Higgins"
output: html_document
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(webexercises)

birthdays <- read.csv("https://raw.githubusercontent.com/higgi13425/medicaldata/master/data-raw/join_data/birthdays.csv") %>% select(-X)

hometowns <- read.csv("https://raw.githubusercontent.com/higgi13425/medicaldata/master/data-raw/join_data/hometowns.csv") %>% select(-X)



```

# Mutating Joins to Combine Data Sources

## What are Joins?

Joins (sometimes called merge operations) are used to join together
different datasets. This can be really helpful for joining together data
from different data sources, like different forms in REDCap
(demographics, vitals, visit 1 data), or the results of a data query
from your electronic medical record, or national or state-level data
available from sources like the CDC or a state health department. By
linking these data you can find new connections, and gain a new
understanding of disease, social determinants of health, or geographic
factors.

## What are Mutating Joins?

These are joins between a base dataset (x) and a new dataset (y) that
add new variables to the base dataset, like a *mutate()* function in
dplyr. Typically you would start with a dataset that includes one row
for each observation, and often one row for each participant.

There are four types of mutating joins:

-   **left join(x,y) ** - all variables from the base dataset (x) are
    retained, and new variables from new dataset (y) that match the
    observations (rows) in dataset x on the key (unique id) variable are
    added.

-   **right join(x,y) ** - all variables from the new dataset (y) are
    retained, and new variables from the base dataset (x) that match the
    observations (rows) in dataset y on the key (unique id) variable are
    added.

-   **inner join(x,y) ** - includes all rows in **both** dataset (x) and
    dataset (y), requiring matching on the key (unique id) variable.

-   **full join(x,y) ** - includes all rows in **either** dataset (x) or
    dataset (y), matching on the key (unique id) variable when matches
    are present, and filling in NAs in missing columns if no match is
    present.

## Let's Start with Left Joins

To get started, let's download two toy datasets - Copy and run the code
chunk below to assign these to birthdays and hometowns.
- Note that you will need to have loaded (library) the tidyverse package to use the pipe.

```{r}
birthdays <- read.csv("https://raw.githubusercontent.com/higgi13425/medicaldata/master/data-raw/join_data/birthdays.csv") %>% select(-X)

hometowns <- read.csv("https://raw.githubusercontent.com/higgi13425/medicaldata/master/data-raw/join_data/hometowns.csv") %>% select(-X)
```

The left join is the most common mutating join, as it allows you to
start with a base dataset (x) and add matching variables from other sources (y).

-   **left join** - all variables from the base dataset (x) are
    retained, and new variables from new dataset (y) that match the
    observations (rows) in dataset x on the key (unique id) variable are
    added.

Let's start with a simple left join with these two small and simple
"toy" datasets about famous physicians. Start by copying and running a
glimpse() function on each dataset from the code chunk below.

```{r glimpse}
glimpse(birthdays)
glimpse(hometowns)
```

We now have two small datasets, with 3 rows each and 2 variables each.
Across the toy datasets, incomplete data are provided on four famous physicians, so
that each dataset is missing one physician from the other dataset.

The 3 main arguments of the left_join() function include

-   **x** (base dataset, or left dataset)

-   **y** (new dataset, or right dataset)

-   **by** (variable to match, known as the key, or unique id - must be
    unique to each observation). The dplyr functions are often smart
    enough to figure the keyid variable out on their own.

## Left Join in Action

The animation below illustrates how a left_join works. The colored
numbers represent the unique key variable, and the gray boxes represent
the unique values from datasets x and y. Notice that observation 4 in
dataset y, because it does not have a match in dataset x, does not end
up in the final dataset.

![left join animation by Garrick Aden-Buie from the `tidyexplain`
repo](https://raw.githubusercontent.com/gadenbuie/tidyexplain/main/images/left-join.gif)

## Left Join in Practice

Let's see how it works. Run the code chunk below. The hometowns dataset
is `x`, or the base/left dataset, and the birthdays dataset is the `y`,
or new/right dataset. Notice that we don't specify a variable for `by`.

```{r}
left_join(x = hometowns, y = birthdays)
```

This produces a new dataset with 3 columns - name, hometown, and date of
birth. The `left_join` function figures out which is the key id variable
to match datasets on its own. You could assign this dataset to a new
object, and add more data from a 3rd, or even a 4th dataset, as long as
each has a matching unique key variable.

::: challenge
## Quick Quiz

Which variable in the left_join above is the key unique variable present
in both datasets that is used to join these two datasets?
`r webexercises::mcq(c(answer = 'name', 'hometown', 'dob'))`

Why does Hippocrates have an NA for date of birth?
`r webexercises::mcq(c('Hippocrates does not appear in the _hometown_ dataset', answer = 'Hippocrates does not appear in the _birthdays_ dataset', 'Hippocrates had a fear of candles', ' Hippocrates always told everyone he was 29'))`
:::

This is great - **left_join()** figured out how to match up the
observations on its own. But sometimes it is not obvious how to match up
observations.

To play it safe, you should specify the matching unique key id variable.
In this case, the common variable to the two datasets is "name". You
specify this with the `by` variable in quotes, as in the below code
chunk

```{r}
left_join(x = hometowns, y = birthdays, by = "name")
```

This gets complicated when the *intended* matching variable does not
quite match.

::: warning
A common "gotcha" when joining datasets is not having an **exact match**
of the intended *by* variable.

Common scenarios include:

-   mismatched spelling or capitalization of the variable name
-   mismatch in the variable data type: integers and numbers **look**
    the same, but if they are not the same data type, the join will
    fail.

It is important to:

-   **glimpse()** both datasets, and check the *name* of the intended
    *by* variable in both datasets.
-   **glimpse()** both datasets, and check the *data type* of the
    intended *by* variable in both datasets.
-   **predict** the number of rows and columns in the resulting dataset
    after the join - if it is way off, something is wrong (a *really*
    important reality check)
:::

## Problem variable names

Sometimes your variable names will be just a bit off. Read in this new
version of hometowns2 by copying and running the code below.

```{r}
hometowns2 <- read.csv("https://raw.githubusercontent.com/higgi13425/medicaldata/master/data-raw/join_data/hometowns2.csv") %>% select(-X)
```

Now write your own left join of hometowns2 and birthdays using the code
below as a starting point. Remember to *glimpse( )* the new dataset
first.

```{r, error=TRUE}
left_join(x = hometowns2, y = birthdays, by = "name")
```

Note that the by variable is not in both datasets, which throws an error. You can fix this in two ways: 

1. Change the variable name in one of the datasets to match the other dataset, then do the join.
2. Use the `by = c("varname_in_x" = "varname_in_y")` syntax to specify the matching variables in each dataset.
Try one of these approaches to get the left join to work.

## Right Join in Action

The animation below illustrates how a right_join works. The colored
numbers represent the unique key variable, and the gray boxes represent
the unique values from datasets x and y. Notice that observation 3 in
dataset x, because it does not have a match in dataset y, does not end
up in the final dataset. In a right join, the y dataset is considered
the base dataset, and the matching observations from the x dataset are
added. This is not very different from a left_join, but occurs in a
different order.

![right join animation by Garrick Aden-Buie from the `tidyexplain`
repo](https://raw.githubusercontent.com/gadenbuie/tidyexplain/main/images/right-join.gif)

## Right Join in Practice

Let's see how it works. Run the code chunk below. The hometowns dataset
is `x`, or the base/left dataset, and the birthdays dataset is the `y`,
or new/right dataset. Notice that we don't specify a variable for `by`.

```{r}
right_join(x = hometowns, y = birthdays)
```

This produces a new dataset with 3 columns - name, hometown, and date of
birth. Notice that this result is **different** from the `left_join`
result above. William Osler appears (without a hometown, as he was born
in the wilds of Ontario), and Hippocrates does not. Which dataset is the
base (x) dataset affects the result of the join.

The `right_join` function figures out which is the key id variable to
match datasets on its own. You could assign this dataset to a new
object, and then stepwise add more data from a 3rd, or even a 4th dataset, as long as
each has a matching unique key variable.

Most people entirely use *left_join()*, rather than *right_join()*, for
the sake of consistency. The main time you might want to use a
*right_join()(*) is when you have a base dataset, and you want to add
some variables from a new dataset, but the new dataset needs some
wrangling and selecting variables first. In this case, you can do your
wrangling of the new dataset, then *right_join()* it to your base
dataset. This will add the new variables to the base dataset.

A simple example is shown below. We want to add the dataset hometowns2
to the dataset *birthdays*, but we first have to remove an extra variable (X), and fix
the "Name" variable name. Then we can pipe the resulting dataset into
the join. But if we pipe it into a *left_join()*, hometowns will by default
become the base (x) dataset, as it comes first. If we want to keep birthdays
as the base (x) dataset, we can use a *right_join()*. This will result in a table in which everyone has a known birthday, but not necessarily a known hometown. 

```{r}
hometowns2 <- read.csv("https://raw.githubusercontent.com/higgi13425/medicaldata/master/data-raw/join_data/hometowns2.csv") %>% 
  select(-X) %>% 
  purrr::set_names(c("name", "hometown"))
hometowns2 %>% 
  right_join(birthdays)
```

## Inner Joins

The inner join only keeps observations for which there is a matching
observation in **both** datasets. Think of the inner_join as the intersection of the two sets. Rows without matches in both datasets will be dropped. This will result in a smaller dataset, but with more complete data.
Here is an animation of what this looks like

![inner join animation by Garrick Aden-Buie from the `tidyexplain`
repo](https://raw.githubusercontent.com/gadenbuie/tidyexplain/main/images/inner-join.gif)

For birthdays and hometowns, what do
you predict the result will be?
Look at the datasets below (run the glimpse code) and take a guess.

```{r}
glimpse(birthdays)
glimpse(hometowns)
```


::: challenge
## Quick Quiz

How many rows will result if you inner_join birthdays with hometowns?
`r webexercises::mcq(c('4', '3', answer = '2', '1'))`

Which physicians will appear in this resulting table?
`r webexercises::mcq(c('Osler and Hippocrates', 'Apgar, Virchow, and Osler', answer = 'Apgar and Virchow', 'Virchow and Hippocrates'))`
:::

## Now Let's take a Look at the result

`r hide("Click to Reveal")`
```{r}
inner_join(birthdays, hometowns)
```
The inner_join requires full matches, so only Virchow and Apgar, who appear in both datasets, appear in the inner_join. 
`r unhide()`

## Full Joins

The full (aka outer) join keeps *all* observations for from **both** datasets, whether or not there is a match. Think of the full_join as the union of the two sets. This will result in a larger dataset but one with more missing data.
Here is an animation of what this looks like

![full join animation by Garrick Aden-Buie from the `tidyexplain`
repo](https://raw.githubusercontent.com/gadenbuie/tidyexplain/main/images/full-join.gif)

For birthdays and hometowns, what do
you predict the result will be?
Look at the datasets below (run the glimpse code) and take a guess.

```{r}
glimpse(birthdays)
glimpse(hometowns)
```


::: challenge
## Quick Quiz

How many rows will result if you full_join birthdays with hometowns?
`r webexercises::mcq(c(answer = '4', '3', '2', '1'))`

Which physicians will appear in this resulting table?
`r webexercises::mcq(c('Osler and Hippocrates', answer = 'Apgar, Osler, Hippocrates and Virchow', 'Apgar, Virchow, and Osler', 'Virchow and Hippocrates'))`
:::

## Now Let's take a Look at the result
`r hide("Click to Reveal")`
```{r}
full_join(birthdays, hometowns)
```
The full_join does not require matches, so all observations from both datasets, appear in the full_join result. 
`r unhide()`

## Complicated Situations

### Repeated Variables
When the same variable name appears in both datasets but is not the key variable, R will append suffixes to the variable names to distinguish them. This can happen when you are joining datasets with similar variables, like height and weight from two different visits. These new variables are named with the original_name.x and original_name.y suffixes. You can change these suffixes with the `suffix` argument in the join function.

### Keeping a limited set of variables
When you need to bring in a key variable from another dataset, but you don't want to bring in all the other variables. You can use the `select()` function from dplyr to select only the key variable and any other variables you want to bring in from the new dataset before doing the join. This keeps your resulting dataset smaller and more manageable.

### There are Some NAs or NANs in the Key variable for matching

If there are missing values in the key variable, these rows will not match during the join, and will result in NAs in the resulting dataset. You may need to clean or impute these missing values before doing the join to ensure accurate matching.
