---
title: "AnalyzeData"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
# load the file with the data you want
# you can also use processed-data-plater.csv and it will work identically
d <- read_csv("data/processed-data-as-is.csv")
```

## Delta Ct

Here I calculate the delta Ct. This could be made into a function. You need to know: 

* Which gene is the reference gene (and how to handle multiple)
* Which columns provide the matching information. I.e. GAPDH value from well E01 is the control for which other wells? In this case, the Sample# and Sample ID columns provide that information. 

```{r}
reference <- d %>% 
  filter(Gene == "GAPDH") %>% 
  mutate(GapdhCt = Cт) %>% 
  select(`Sample#`, `Sample ID`, GapdhCt)

d <- inner_join(d, reference) %>% 
  mutate(dCt = GapdhCt - Cт)

```

## Plot dCt
Recreate plot from the Excel file
```{r}
ggplot(d, aes(x = `Sample ID`, y = dCt)) + geom_point() + 
  facet_wrap(~Gene) + theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

## Average across biological replicates
Here I average the dCt across the biological replicates. To make this a function you need to know which columns contain the Gene and matching information. In this case, it's the Gene and the Sample ID columns.

```{r}
d <- d %>% 
  group_by(Gene, `Sample ID`) %>% 
  summarise(AverageDCt = mean(dCt))
```

## ddCt and fold change
Here I calculate ddCt and fold change. To make this a function, you need to know which columns contain the information allowing you to identify which wells are the baseline wells, i.e. the wells to which the ddCt should be calculated.

```{r}
baseline <- d %>% 
  filter(`Sample ID` == "Un") %>% 
  mutate(BaselineDCt = AverageDCt) %>% 
  select(Gene, BaselineDCt)

d <- inner_join(d, baseline) %>% 
  mutate(ddCt = AverageDCt - BaselineDCt, FoldChange = 2^ddCt)

ggplot(d, aes(x = `Sample ID`, y = FoldChange)) + geom_point() + 
  facet_wrap(~Gene) + theme(axis.text.x = element_text(angle = 45, hjust = 1))

```