---
title: "Lasso"
author: "Peter Higgins"
date: "2024-11-21"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(glmnet)
```

## Define

```{r}
# base R dialect
# y <- mtcars$hp 
# # x <- data.matrix(mtcars[, c('mpg', 'wt', 'drat', 'qsec')])

# tidyverse dialect
y <- mtcars |> select(hp) |> pull()

x <- mtcars |> select(mpg, wt, drat, qsec) |> data.matrix()
```


# Model

```{r cars}
cv_model <- cv.glmnet(x, y, alpha = 1) 

best_lambda <- cv_model$lambda.min 

best_lambda

plot(cv_model) 

best_model <- glmnet(x, y, alpha = 1, lambda = best_lambda) 
coef(best_model)

best_model |> 
  broom::tidy() 

best_model |> 
  broom::glance() 
```

## Predictions


```{r predict}
new = matrix(c(24, 2.5, 3.5, 18.5), nrow=1, ncol=4) 

predict(best_model, s = best_lambda, newx = new) 
```

Calculate r squared

```{r}
y_predicted <- predict(best_model, s = best_lambda, newx = x) 

# SST - sum of squares total - diffs between values and mean, squared and sum
sst <- sum((y - mean(y))^2) 
# SSE = sum of squared estimate of errors
sse <- sum((y_predicted - y)^2) 

#find R-Squared 
rsq <- 1 - sse/sst 
rsq

```

## Indomethacin RCT
```{r}
indo_rct <- medicaldata::indo_rct
```

### define response variable, matrix
```{r}
y <- parse_number(as.character(indo_rct$outcome))

x <- data.matrix(indo_rct[, c('site', 'age', 'risk','gender',
                'sod', 'recpanc', "psphinc", 'precut', 'difcan',
                'pneudil', 'amp', 'paninj', 'acinar', 'brush',
                'asa81', 'asa325', 'asa', 'prophystent', 'therastent',
                'pdstent', 'sodsom', 'bsphinc', 'bstent', 'chole',
                'pbmal', 'train', 'status', 'type', 'rx')])
```

### Modeling
```{r}
cv_model <- cv.glmnet(x, y, alpha = 1) 

best_lambda <- cv_model$lambda.min
best_lambda

plot(cv_model)
```

### Show Coefficients

```{r}
best_model <- glmnet(x, y, alpha = 1, lambda = best_lambda)
coef(best_model)

best_model |> 
  broom::tidy() 
```

# Example frm Australia
```{r}
# install.packages('faraway')
library(faraway)
set.seed(1233)
data("fat")
head(fat)
```

# set up modeling
```{r}
Y <- fat$brozek

X <- model.matrix(brozek ~ age + weight + height + neck + chest + abdom + hip + thigh + knee + ankle + biceps + forearm + wrist, data=fat)[,-1] 

```

# optimize lambda

```{r}
cv.lambda.lasso <- cv.glmnet(x=X, y=Y, alpha = 1) 

plot(cv.lambda.lasso) 

cv.lambda.lasso
```

# Now Model with optimal lambda
```{r}
l.lasso.min <- cv.lambda.lasso$lambda.min 

lasso.model <- glmnet(x=X, y=Y, alpha = 1, lambda = l.lasso.min)

lasso.model$beta
```

### Compare to OLS
```{r}
ols.model <- glm(brozek ~ age + weight + height + adipos + neck + chest + abdom + hip + thigh + knee + ankle + biceps + forearm + wrist, data=fat) 
summary(ols.model)
```

## Lasso in Logistic
Breast cancer 0/1 ~ microscopic findings in biopsy histopath, lots of correlations

```{r}
library(corrplot) #plots the correlation heatmap
library(pROC)
set.seed(122) 

dados<-read.csv("https://www.dropbox.com/s/vp44yozebx5xgok/bdiag.csv?dl=1") 
correlations <- cor(dados[,-c(1,2)]) 
#correlations among all the predictors #excluding ID and DIAGNOSIS, columns 1&2 
corrplot(correlations) #plots the correlations – some vars Highly Corr

```

## See problems with correlations
```{r}
summary(glm(diagnosis == "M" ~ radius_mean , family=binomial, data=dados )) 
summary(glm(diagnosis == "M" ~ radius_mean + perimeter_mean, family=binomial, data=dados ))

```

### Model fit with Lasso
```{r}
X <-  model.matrix(diagnosis ~  . | pt_id, data=dados[,-1])[,-1]  

Y <- dados[,"diagnosis"]=="M"       

# find optimal lambda
cv.model<- cv.glmnet(x=X, y=Y,
                     family = "binomial",
                     alpha=1) #alpha=1 is lasso 
plot(cv.model)
l.min <- cv.model$lambda.min

```

### Fit model with optimal lambda
```{r}
lasso.model <- glmnet(x=X, y=Y, 
                      family = "binomial",	
                      alpha=1, lambda = l.min) 
lasso.model$beta

lasso.model |> 
  broom::tidy() 
```

### Assess the model
```{r}
assess.glmnet(lasso.model, newx=X, newy = Y)
#in this case, we are evaluating the model newx = X, 
#in the same data used to fit the model newy = Y ) 
plot(roc.glmnet(lasso.model, newx = X, newy = Y ), 	type="l")

```

