---
title: "GGRidges"
author: "Peter Higgins"
date: "5/27/2021"
output: html_document
bibliography: references.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(ggridges)
library(medicaldata)
library(janitor)

cmv <- medicaldata::cytomegalovirus 

covid <- medicaldata::covid_testing %>% 
  filter(patient_class != "not applicable") %>% 
  filter(!is.na(patient_class)) %>% 
  filter(patient_class != "day surgery") %>% 
  filter(patient_class != "admit after surgery-ip") %>% 
  filter(patient_class != "admit after surgery-obs")
```

## Ridgeline Plots

Ridgeline plots are a nice way to show the distribution of a continuous variable, and compare distributions across several categories

```{r plot}
cmv %>% 
ggplot(aes(x = time.to.transplant, y = diagnosis)) + geom_density_ridges(scale = 0.9) 

```

## Including Plots

You can also embed plots, for example:

```{r plot2}
cmv %>% 
ggplot(aes(x = time.to.agvhd, y = diagnosis,
           fill= stat(x))) + 
  geom_density_ridges_gradient(scale = 0.9) +
  scale_fill_viridis_c(name = "Days", option = "A") +
  labs(title = "Time to AGvHD in Days")
```

## Including Points

You can also add data points, for example:

```{r plot3}
cmv %>% 
ggplot(aes(x = time.to.cmv, y = diagnosis,
           fill= stat(x))) + 
  geom_density_ridges_gradient(scale = 0.5,
        jittered_points = TRUE,
        position = "raincloud",
        alpha = 0.5) +
  scale_fill_viridis_c(name = "Days", option = "B") +
  labs(title = "Time to CMV in Days") +
  theme_ridges()
```

## Including Points

You can also add data points, for example:

```{r plot4}
cmv %>% 
ggplot(aes(x = CD8.dose, y = diagnosis,
           fill= stat(x))) + 
  geom_density_ridges_gradient(scale = 0.9) +
  scale_fill_viridis_c(name = "Days", option = "C") +
  labs(title = "CD8 Dose") +
  theme_ridges()
```

## Including Points

You can also add data points, for example:

```{r plot5}
cmv %>% 
ggplot(aes(x = CD3.dose, y = diagnosis,
           fill= stat(x))) + 
  geom_density_ridges_gradient(scale = 0.7,
        jittered_points = TRUE,
        position = "raincloud") +
  scale_fill_viridis_c(name = "Days", option = "D") +
  labs(title = "CD3 Dose") +
  theme_ridges()
```

## Including Points

You can also add data points, for example:

```{r plot6}
cmv %>% 
ggplot(aes(x = TNC.dose, y = diagnosis,
           fill= stat(x))) + 
  geom_density_ridges_gradient(scale = 2,
        jittered_points = TRUE,
        position = "points_sina",
        alpha = 0.7) +
  scale_fill_viridis_c(name = "Days", option = "E") +
  labs(title = "TNC Dose") +
  theme_ridges()
```

## 

[@stidham2012]
